---
title: "Get Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{tidyverse, patchwork}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

## Installation

Please see the [Installation and Setup](https://schisslergroup.github.io/bigsimr/articles/install-bigsimr.html) page for getting `bigsimr` up and running. `bigsimr` requires a working version of Python that can be found in the path variable. It is also required that the version of python has `numpy` and `jax` installed.

## Usage

```{r, eval=FALSE}
library(bigsimr)
library(tidyverse)
library(patchwork)
```

We're going to show the basic use and syntax of `bigsimr` by using the New York air quality data set (`airquality`) included in the R `datasets` package. We will focus specifically on the temperature (degrees Fahrenheit) and ozone level (parts per billion).

```{r, eval=FALSE}
df <- airquality %>%
  select(Ozone, Temp) %>%
  drop_na()

glimpse(df)
```

Let's look at the joint distribution of the Ozone and Temperature

```{r, eval=FALSE}
p0 <- ggplot(df, aes(Temp, Ozone)) +
  geom_point(size = 1) +
  theme(legend.position = "none") +
  labs(x = "Temperature")

pTemp <- ggplot(df, aes(Temp)) + 
  geom_density() +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())

pOzone <- ggplot(df, aes(Ozone)) + 
  geom_density() + 
  theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank()) +
  coord_flip() 

pTemp + plot_spacer() + p0 + pOzone + 
  plot_layout(widths = c(3,1), heights = c(1, 3))
```

We can see that not all margins are normally distributed; the ozone level is highly skewed. Though we don't know the true distribution of ozone levels, we can go forward assuming that it is log-normally distributed.

To simulate observations from this joint distribution, we need to estimate the correlation and the marginal parameters.

### Estimating Correlation

To estimate the correlation, we can use `R`'s function `cor` or we can use the convenience function `bigsimr::cor_fast` which can estimate Pearson, Spearman, or Kendall correlation using the fastest methods available from other packages.

```{r, eval=FALSE}
(rho <- cor_fast(df, method = "pearson"))
```

### Defining Marginal Distributions

Next we can estimate the marginal parameters. Assuming that `Temperature` is normally distributed, it has parameters

```{r, eval=FALSE}
df %>% select(Temp) %>% summarise_all(.funs = c(mean = mean, sd = sd))
```

and assuming that `Ozone` is log-normally distributed, it has parameters

```{r, eval=FALSE}
mle_mean <- function(x) mean(log(x))
mle_sd <- function(x) sqrt(mean( (log(x) - mean(log(x)))^2 ))

df %>% select(Ozone) %>% 
  summarise_all(.funs = c(meanlog = mle_mean, sdlog = mle_sd))
```

Finally we take the parameters and put them into an `alist` of margins.

```{r, eval=FALSE}
margins <- alist(
  qlnorm(meanlog = 3.42, sdlog = 0.862),
  qnorm(mean = 77.9, sd = 9.49)
)
```

Right now the marginal parameters must be filled in manually with the quantile function of the distribution. If the margins are homogeneous, then some automation can be implemented.

### Correlation Bounds

Given a list of margins, the theoretical lower and upper correlation coefficients can be estimated.

```{r, eval=FALSE}
cor_bounds(margins = margins, type = "pearson")
```

Since our estimated correlation is within the theoretical bounds (assuming the distributions are correct), we should be able to achieve a similar correlation in the simulated data.

### Simulating Multivariate Distributions

Let's now simulate 10,000 observations from the joint distribution using `bigsimr::rvec`

```{r, eval=FALSE}
x <- rvec(10000, rho, margins, "pearson")

df_sim <- as.data.frame(x)
colnames(df_sim) <- colnames(df)
```

and visualize the joint distribution

```{r, eval=FALSE}
p1 <- df_sim %>%
  ggplot(aes(Temp, Ozone)) +
  geom_density_2d_filled() +
  theme(legend.position = "none") +
  labs(x = "Temperature", y = "Ozone") +
  scale_y_continuous(limits = c(0, max(df$Ozone))) +
  scale_x_continuous(limits = range(df$Temp))

p1Temp <- ggplot(df_sim, aes(Temp)) + 
  geom_density(alpha = 0.5, fill = "lightseagreen") +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())

p1Ozone <- ggplot(df_sim, aes(Ozone)) + 
  geom_density(alpha = 0.5, fill = "lightseagreen") + 
  theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank()) +
  coord_flip() 

p1Temp + plot_spacer() + p1 + p1Ozone + 
  plot_layout(widths = c(3,1), heights = c(1, 3))
```

## Comparison to Uncorrelated Samples

We can compare the bivariate distribution above to one where no correlation is taken into account.

```{r, eval=FALSE}
df_sim2 <- data.frame(
  Ozone = rlnorm(10000, mle_mean(df$Ozone), mle_sd(df$Ozone)),
  Temp = rnorm(10000, mean(df$Temp), sd(df$Temp))
)
```

```{r, echo=FALSE, warning=FALSE, eval=FALSE}
p2 <- df_sim2 %>%
  ggplot(aes(Temp, Ozone)) +
  geom_density_2d_filled() +
  theme(legend.position = "none") +
  labs(x = "Temperature", y = "Ozone") +
  scale_y_continuous(limits = c(0, max(df$Ozone))) +
  scale_x_continuous(limits = range(df$Temp))

p2Temp <- ggplot(df_sim2, aes(Temp)) + 
  geom_density(alpha = 0.5, fill = "lightseagreen") +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())

p2Ozone <- ggplot(df_sim2, aes(Ozone)) + 
  geom_density(alpha = 0.5, fill = "lightseagreen") + 
  theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank()) +
  coord_flip() 

p2Temp + plot_spacer() + p2 + p2Ozone + 
  plot_layout(widths = c(3,1), heights = c(1, 3))
```

Notice that this uncorrelated bivariate distribution has the same marginal distributions as the correlated example above.
