---
title: "Using bigsimr::rvec"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{using-rvec}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

reticulate::use_condaenv("py37")
devtools::load_all()
```

```{r setup, eval=FALSE}
 # Optional step if your python binary is in a miniconda environment
reticulate::use_condaenv("py37")
library(bigsimr)
```

# Installation

This package requires a working python distribution (typically via a conda environment). As of right now, we cannot automatically create the necessary environment for you, but you can do so by installing Anaconda or Miniconda, and then installing `jax` and `jaxlib` ([see here](https://github.com/google/jax#installation)). There is an option for install `jax` with GPU support. This is recommended if possible, otherwise `bigsimr` will simply throw a warning stating that it has fallen back onto the CPU.

# Basics

This package operates on the principle of Gaussian copulas, for which you will need a list of target marginal distributions and a correlation structure between the margins.

The main function is `bigsimr::rvec` which mirrors some of the conventions found in the base `stats` package like `stats::rnorm`. I.e. the common convention for generating a random vector is to first specify the number of samples followed by the parameters for the distribution.

```{r}
# Generating samples from a normal distribution
# We need three things: 
#   1) The number of samples to generate
#   2) The location of the normal distribution (mean)
#   3) The scale of the normal distribution (standard deviation)

rnorm(n = 16, mean = 5, sd = 1.5)
```

The 'r' in front of 'norm' is standard for 'random', as in generate random samples from the normal distribution. Other examples are `rexp` (exponential), `rnbinom` (negative binomial), and `rt` (Student t).

One thing to familiarize yourself with is the specific parameterization of certain R distributions. `stats::rexp` uses a rate parameter while others may use the scale parameter, $1 / rate$. Our package uses the default parameterizations found in R, so be sure to check that you have the correct parameterization beforehand.

# Building a Multivariate Distribution

As stated earlier, to generate multivariate data, we need a list of marginals (and their parameters), and a correlation structure (matrix). The marginal distributions can be built up as a list of lists, where each sublist contains the information for the target distribution.

```{r}
margins = list(
  list("norm", mean = 3.14, sd = 0.1),
  list("beta", shape1 = 1, shape2 = 4),
  list("nbinom", size = 10, prob = 0.75)
)
```

The things to point out here are that in each sublist (marginal), the first item is an unnamed character string with the R name of the distribution *without a letter prefix*. E.g. instead of `rnorm`, we pass in just `"norm"`. The second thing to note is that the remaining items are *named* arguments that go along with the distribution. A full list of built-in distributions is found in the appendix.

The next step is to define a correlation structure for the multivariate distribution. This correlation matrix can either come from observed data, or we can set it ourselves, or we can generate a random correlation matrix via `bigsimr::rcor`. Let's create a simple correlation matrix where all off-diagonal elements are 0.5. Since we have 3 marginals, we need a $3\times 3$ matrix.

```{r}
rho <- matrix(0.5, nrow = 3, ncol = 3)
diag(rho) <- 1.0
rho
```

Finally we can generate a random vector with our specified marginals and correlation structure. The last argument, `type`, is looking to know what kind of correlation matrix it is receiving. Right now it can handle Pearson, Spearman, or Kendall.

```{r}
x <- rvec(10, rho = rho, params = margins, type = "pearson")
```

On my machine, there is no dedicated GPU, so I would see the following warning message once per session.

```{r, echo=FALSE}
warning("warning.warn('No GPU/TPU found, falling back to CPU.')")
```

Taking a look at our random vector, we see that it hs 10 rows and 3 columns, one column for each marginal.

```{r}
x
```

We can simulate many more samples and then check the histogram of each margin, as well as the estimated correlation between the columns.

```{r, fig.width=7}
x = rvec(10000, rho = rho, params = margins, type = "pearson")

par(mfrow=c(1,3))
hist(x[,1], breaks = 30, xlab = "", main = "Normal")
hist(x[,2], breaks = 30, xlab = "", main = "Beta")
hist(x[,3], breaks = 30, xlab = "", main = "Negative Binomial")
```

```{r}
cor(x)
```

We can see that even wih 10,000 samples, the estimated correlation of the simulated data is not exactly the same as the target correlation. This can be explained by the fact that some correlations are simply not possible due to the discrete nature of certain distributions. Another possibility is that the copula algorith is biased and needs correction. 

# Appendix

```{r, eval=FALSE}
all_dists <- list(
  list(dist = "beta", shape1, shape2),
  list(dist = "binom", size, prob),
  list(dist = "cauchy", location, scale),
  list(dist = "chisq", df),
  list(dist = "exp", rate),
  list(dist = "f", df1, df2),
  list(dist = "gamma", shape, rate),
  list(dist = "geom", prob),
  list(dist = "hyper", m, n, k),
  list(dist = "logis", location, scale),
  list(dist = "lnorm", meanlog, sdlog),
  list(dist = "nbinom", size, prob),
  list(dist = "norm", mean, sd),
  list(dist = "pois", lambda),
  list(dist = "t", df),
  list(dist = "unif", min, max),
  list(dist = "weibull", shape, scale),
  list(dist = "wilcox", m, n),
  list(dist = "signrank", n)
)
```
